---
title: "Getting started with joinery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with joinery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
set.seed(1)
```

## The problem

You have two tables of people. They are the same people, typed by different
hands on different days. One table writes `Prof. Dr. Müller`, the other writes
`Mueller`. One has `Bahnhofstraße 67a`, the other `Bahnhofstr. 67 A`. First names
collapse to initials. Middle names appear and disappear. Districts are spelled
the same on both sides because nobody mistypes where they live.

An exact join finds none of these. Edit distance does a little better, but it
trips on the things that matter: it counts `Müller` and `Mueller` as two edits
apart, even though they are the same name, while rating genuinely different
surnames that sit one letter apart, like `Bauer` and `Mauer`, as near. It has no idea that a rare surname is strong evidence and a common one is
weak. It compares how strings *look*, when what you want to compare is the
*information* two records share.

joinery takes the other route. It cuts each field into tokens, weights each
token by how rare it is, and scores a pair by the rare tokens they have in
common. This vignette walks the whole path on a pair of built-in tables, and
then does something most linkage tutorials cannot: it scores the result against
a known answer key.

## How joinery thinks about a match

Take one real pair from the data we are about to load: a register row reading `Amelie Hofmann, Konrad-Adenauer-Straße 2,
Ronnenberg` and a listing reading `A. Hofmann, Konrad-Adenauer-Straße 2,
Ronnenberg`.

1. **Tokens.** Each field is cut into tokens, with the spelling smoothed first
   (lowercased, accents stripped) so that `Amelie` and `amelie` are the same
   token. The listing's lone initial `A.` carries almost no information and will
   not match the register's full `amelie`. How exactly each field is cut is the
   strategy you write in a moment.
2. **Rarity.** A token is worth more when it is rare. `Hofmann` is a common
   surname, so sharing it says something but not much. `Ronnenberg` is a rare
   place, so sharing it says a lot. This single idea is what separates joinery
   from edit-distance matching.
3. **Overlap, not appearance.** The two records are compared by the rare tokens
   they share, not by how their strings line up. The missing first name costs
   nothing, because the score is built from what the records have in common, not
   from what is absent.
4. **Score and threshold.** Each shared token contributes its share of its
   column's weight. The contributions add up to a score between 0 and 1. A pair
   is kept when the score clears the **threshold**.
5. **Entities.** Records that link, directly or through a chain of links, form
   an **entity**: a duplicate group within one table, or a matched cluster
   across two.

The pair above scores exactly `0.80`, and we will see later, token by token,
exactly where that score came from. That per-token receipt is `explain_match()`,
and it is the whole reason to prefer a transparent matcher: you can always ask
why a pair scored what it did.

## 1. Look at the data

```{r data}
library(joinery)
library(dplyr)

data(base_example)
data(target_example)

glimpse(base_example)
```

joinery works directly with the data frames and tibbles you already have;
there is nothing to convert. We use a few `dplyr` verbs below only to slice the
results, not to feed joinery.

`base_example` is 3,300 person records. The last 300 are deliberate
near-duplicates of earlier rows: the same people, re-entered with initials,
added titles, dropped middle names, and house-number noise. That is the
duplicate-detection case.

```{r target}
glimpse(target_example)
```

`target_example` is 3,000 records. About 80% are distorted copies of
`base_example` people; the rest are genuinely new. The first column,
`actual_link`, is the answer key: for every copied row it holds the true
`id_base` it came from. That is unusual for a linkage tutorial and we will lean
on it in section 7.

Inspect one known pair. Target row `T0006` carries `actual_link = "B0006"`:

```{r pair}
target_example |>
  filter(id_target == "T0006") |>
  select(id_target, Vorname, Nachname, Strasse, Ort)

base_example |>
  filter(id_base == "B0006") |>
  select(id_base, Vorname, Nachname, Strasse, Ort)
```

Same person, `Amelie` shortened to `A.`. An exact join on name misses it.

`Kreis` (the administrative district) is the one field the noise leaves alone,
so it makes a natural **blocking** key: only compare records that sit in the
same district. That turns a 3,300 by 3,000 comparison into a handful of small
ones.

## 2. Declare a strategy

Think of a strategy as a jig, a woodworker's template that guides the same
cut every time. You set it up once, then run it over every table. It says how
to turn each column into tokens, how to block, and where to set the threshold.
It runs nothing by itself.

```{r strategy}
strat <- search_strategy(
  Nachname   ~ normalize_text() + word_tokens(min_nchar = 3),
  Vorname    ~ normalize_text() + word_tokens(min_nchar = 3),
  Strasse    ~ normalize_street(lang = "de") + word_tokens(min_nchar = 3),
  Hausnummer ~ numeric_tokens,
  Ort        ~ normalize_text(),
  block_by   = "Kreis",
  threshold  = 0.8
)

strat
```

### Preparers: how a column becomes tokens

Each formula reads `column ~ preparer1 + preparer2 + ...`: a small pipeline,
run left to right. The early steps smooth the text; the last step cuts it into
tokens. The four preparers used above:

- `normalize_text()` lowercases and strips accents, so casing and diacritics
  stop mattering.
- `normalize_street(lang = "de")` expands German street abbreviations
  (`str.` becomes `straße`) before the text is cut.
- `word_tokens(min_nchar = 3)` splits text into words and drops anything shorter
  than three characters, so a lone initial like `A.` falls out while `Amelie`
  stays.
- `numeric_tokens` keeps only the digit runs. It takes no arguments, so it is
  written bare, without the `()`.

So `Ort ~ normalize_text()` smooths the town name but never splits it, keeping it
as one token, while `Nachname ~ normalize_text() + word_tokens(min_nchar = 3)`
smooths *and* splits. joinery ships many more preparers (phonetic encoders such
as `as_metaphone()`, n-grams, stopword filters); the reference index lists them
all.

### The other arguments

The formulas are the only required part. The rest are tuning knobs with sensible
defaults:

| Argument | What it does | If you omit it |
|---|---|---|
| `block_by` | only compare records that share this column's value | no blocking: every record is compared with every other (fine for small tables, costly for large ones) |
| `threshold` | the lowest score a pair can have and still be kept | defaults to `0.9` |
| `weights` | a named vector to make some columns count for more | every column counts equally |
| `rarity` | how a token's rarity is measured | `"inverse_freq"`, where rarer tokens score higher |
| `max_candidates` | maximum candidate matches kept per record; only the top-scoring N are returned | no limit |
| `smoothing` | transforms rIP scores before aggregation, redistributing weight across rare and common tokens | identity (no transformation) |
| `feedback_strength` | penalises a pair when the rare tokens of one record only partially appear in the other | `0` (disabled) |

Here `block_by = "Kreis"` restricts every comparison to within a district, and
`threshold = 0.8` loosens the default slightly. The three bottom rows in the
table appear in the strategy print-out but are rarely needed: `max_candidates`,
`smoothing`, and `feedback_strength` all default to "none" and can be left alone
for most linkage tasks. There are further levers for large runs (`min_rarity`,
`max_token_df`, and a fan-out guard); their defaults are safe, and
`?search_strategy` documents them. This strategy is the one block of new syntax
in the package; the rest are verbs that consume it.

## 3. Will it work? (check before you match)

You do not have to run a match to find out whether a strategy is sound. Start by
inspecting one column's tokens:

```{r inspect}
head(inspect_tokens(base_example, "id_base", strat, Vorname), 8)
```

Then ask for a pre-match health check:

```{r audit}
audit_strategy(base_example, "id_base", strat)
```

`audit_strategy()` reports, per column, how many distinct tokens there are and
how rare they run, plus the block layout and an estimate of how many comparisons
the match will cost. Here the blocking cuts the work to about 309,000
comparisons instead of the ten million a full cross would need. If a column were
all boilerplate (no rare tokens) or a block were so large it would make the comparison count unmanageable,
this is where you would see it, before paying for the match.

When no blocking key is available, `max_candidates` offers a softer control:
setting it to, say, `3` keeps only the three highest-scoring candidates per
record, capping the output without changing which pairs are evaluated. It does
not reduce computation the way blocking does, but it prevents a large result
table when the threshold alone is too loose.

If a single hyper-common token (a frequent house number, say) were fanning a
block out, `rarity_distribution()` would show it and let you set the `min_rarity`
or `max_token_df` levers. We do not need it here.

## 4. Deduplicate the base table

Deduplication is just matching a table against itself with the same strategy.

```{r dedup}
dups <- detect_duplicates(base_example, id = "id_base", strategy = strat)

dups |>
  select(duplicate_group, id, score, rank) |>
  head()
```

`deduplicate_table()` collapses each duplicate group to a single record:

```{r dedup-collapse}
base_clean <- deduplicate_table(base_example, dups, id = "id_base")

nrow(base_example) - nrow(base_clean)
```

We planted exactly 300 duplicates (the last 300 rows), and the dedup recovers
282 of them. The rest are the 18 where the noise was heavy enough to drop the
pair below `0.8`, which is exactly the precision/recall trade-off we look at
next.

## 5. Search across tables

Now link the cleaned base table to the target table:

```{r search}
matches <- search_candidates(
  base_clean,
  target_example,
  base_id   = "id_base",
  target_id = "id_target",
  strategy  = strat
)

matches |>
  select(match_id, score, source, id, Nachname, rank) |>
  head()
```

Each `match_id` groups the two sides of one candidate pair: a `base` row and a
`target` row. The `score` is the shared-rarity total; `rank` orders competing
candidates for the same record.

## 6. Did it work, and why this pair?

Every match raises two questions: whether the result held together, and why a
given pair scored what it did.

```{r summarise}
summarise_matches(matches, threshold = 0.8)
```

The overview shows the score distribution and flags how many pairs sit close to
the threshold. For *why this pair*, ask for the receipt. Take the
`Amelie`/`A. Hofmann` pair from the start: find its `match_id`, then explain it.

```{r explain}
mid <- matches |> filter(id == "T0006") |> pull(match_id) |> first()

receipt <- explain_match(
  matches, strat,
  base      = base_clean,
  id        = "id_base",
  target    = target_example,
  target_id = "id_target",
  match_id  = mid
)

receipt
```

This is the pair introduced in section 1. It scores exactly `0.80`.
The first name contributed nothing: the initial `A.` was shorter than the
`min_nchar = 3` cutoff and fell out of the token set. Nachname, Strasse,
Hausnummer, and Ort each contributed `0.2`, because the weights are equal and
each of those columns returned at least one shared token.

The per-token receipt shows where the score came from and where it did not.
Plotting it at token resolution makes the variation visible even when column
totals are uniform:

```{r explain-plot, fig.alt = "Horizontal bar chart of per-token contributions to the match score, coloured by column"}
token_contribution_plot(receipt)
```

Within Strasse, `konrad` and `adenauer` each contribute most of that column's
0.2 share, while `strasse` (a word that appears on nearly every street) earns
almost no rarity weight and contributes little despite being shared. Nothing
about the score is hidden: every token shows its rarity, its rIP, and its
contribution.

## 7. Score against the answer key

Because `target_example$actual_link` is ground truth, we can measure the match
instead of trusting it. For each candidate pair, compare the base id we picked
to the true link:

```{r scoring}
pred <- matches |>
  group_by(match_id) |>
  summarise(
    base_id = id[source == "base"][1],
    truth   = actual_link[source == "target"][1],
    .groups = "drop"
  ) |>
  mutate(correct = base_id == truth)

recoverable <- sum(target_example$actual_link %in% base_example$id_base)

c(
  pairs     = nrow(pred),
  precision = round(mean(pred$correct, na.rm = TRUE), 3),
  recall    = round(sum(pred$correct, na.rm = TRUE) / recoverable, 3)
)
```

At `threshold = 0.8` the matches are all correct (precision `1.0`) and recover
three quarters of the recoverable links (recall `0.75`). The threshold is the
dial between those two. The code below sweeps it across four values:

```{r sweep}
sweep <- bind_rows(lapply(c(0.6, 0.7, 0.8, 0.9), function(th) {
  st <- strat
  st@threshold <- th
  m <- search_candidates(base_clean, target_example,
                         base_id = "id_base", target_id = "id_target",
                         strategy = st)
  p <- m |>
    group_by(match_id) |>
    summarise(base_id = id[source == "base"][1],
              truth   = actual_link[source == "target"][1],
              .groups = "drop") |>
    mutate(correct = base_id == truth)
  tibble(
    threshold = th,
    pairs     = nrow(p),
    precision = round(mean(p$correct, na.rm = TRUE), 3),
    recall    = round(sum(p$correct, na.rm = TRUE) / recoverable, 3)
  )
}))

sweep
```

Dropping to `0.6` lifts recall from `0.75` to `0.95`, at the cost of a few false
links (precision `0.94`). Raising to `0.9` keeps precision perfect but throws
away two thirds of the true matches. There is no single right answer; the right
threshold depends on whether a missed link or a wrong link costs you more. When
you need to do better than a single dial, a trained false-positive filter
(`calibrate_matches()`) learns the boundary from labelled pairs.

## 8. Multistage matching

One pass rarely catches everything. The records a match did not touch are its
residual:

```{r residuals}
unmatched_base   <- extract_unmatched(base_clean, "id_base", matches)
unmatched_target <- extract_unmatched(target_example, "id_target", matches)

nrow(unmatched_base)
nrow(unmatched_target)
```

`multi_stage_search()` handles these residuals automatically; you do not need to
extract and pass them yourself. The calls below show how.

### Exact matching as a first gate

The standard approach is to layer passes: a cheap **exact** stage first, then a
tolerant fuzzy stage only on what is left. The layering logic is that the exact
stage clears the easy cases (pairs where one record's token set is fully
contained in the other's) before the heavier scoring starts on the harder
residual.

An `exact_strategy()` matches pairs where one record's token set for each column
is fully contained in the other's. The score is always `1.0`; there is no
rarity weighting and no threshold to tune. It is fast and produces zero false
positives, so it is a reliable first gate. It takes the same column formulas as
a `search_strategy()`, but weights, `min_rarity`, and thresholds are ignored:
the match criterion is binary containment.

### Composing stages

`multi_stage_search()` composes the stages in one call: it runs the exact
strategy, extracts the residual, runs the fuzzy strategy on what is left, and
merges the results into a single entity ledger.

```{r staged}
staged <- multi_stage_search(
  base_clean, target_example,
  base_id   = "id_base",
  target_id = "id_target",
  strategies = list(
    exact = exact_strategy(
      Nachname ~ normalize_text() + word_tokens(min_nchar = 3),
      Vorname  ~ normalize_text() + word_tokens(min_nchar = 3),
      Ort      ~ normalize_text(),
      block_by = "Kreis"
    ),
    fuzzy = strat
  )
)

table(staged$stage)
```

The stage labels come from the names given in the `strategies` list — here
`"exact"` and `"fuzzy"` — showing how many records each pass placed. Each row
in the result ties a record to the entity it landed in and is tagged with the
stage that placed it. Records claimed by the exact stage never enter the fuzzy
stage, so the fuzzy scorer works on a smaller and harder problem. You can add
as many stages as needed, each with a progressively looser strategy operating
on the residual of the previous, and the ledger tracks which stage placed each
record.

For deduplicating a single table the same way, use `multi_stage_dedup()`; it
runs the same staged logic and resolves connected components at the end.

## 9. Where to look next

You now have the spine: declare a strategy, check it, dedup, search, score,
stage.

- `sample_matches()` and `export_for_labelling()` build a labelling set from a
  match result.
- `fit_filter()` and `calibrate_matches()` train a false-positive filter on
  those labels, for when one threshold is not enough.
- `plan_strategy()` helps choose a blocking key on a dataset you do not know yet.
- The same verbs run on a DuckDB connection when the data is too large for
  memory.

A planned feature article will walk the advanced joins (region-free movers,
phonetic encoders, containment, the fan-out guard) on the `workshop_register` /
`workshop_listings` tables, where each feature has a planted case that
measurably wins when you switch it on.
