---
title: "Working at scale with DuckDB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working at scale with DuckDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

Every other article in this set works on a table that fits in memory. That is the
right tool for a few hundred thousand records, and most jobs never grow past it.
But some do. A business directory published every year for two decades is tens of
millions of listings, and the moment the table no longer fits in RAM the
in-memory backend stops being an option. The matching problem has not changed.
What changed is the bench it sits on.

joinery's answer is that the verbs do not care where the table lives. Hand them a
DuckDB table instead of a data frame and the same code runs: the database holds
the records, does the large joins on disk, and hands back only what you ask for.
You write the strategy once and it runs on either backend, so you can prototype on
a small in-memory slice and run the full corpus through DuckDB without rewriting a
line of matching logic.

This article shows that swap on a small panel you can actually run, then explains
the one new object you need at scale and closes with what the same code did on a
real corpus of 51 million listings.

```{r load}
library(joinery)
library(duckdb)
library(DBI)
library(dplyr)
library(data.table)
```

## The same job, a bigger bench

We use `workshop_panel`, the multi-year panel of woodworking workshops from the
[matching across years and sources](staged.html) article. That article explains
*why* the strategies below are shaped the way they are: an exact pass for the
clean rows, a fuzzy pass for drifting names, a mover pass that blocks on a rare
name token to reach across relocations. Here the strategies are a given. The
subject is the backend underneath them.

The panel is small enough to run in seconds, which is the point: it lets you watch
the backend swap with your own eyes before you trust it on something you cannot
hold in memory.

```{r data}
c(rows = nrow(workshop_panel),
  workshops = uniqueN(workshop_panel$true_entity),
  years = uniqueN(workshop_panel$year))
```

## Putting the table in DuckDB

A DuckDB table is just a connection plus a lazy reference to a named table. You
write your data in once, then point `tbl()` at it. Everything downstream treats
that reference exactly as it would a data frame.

```{r connect}
con <- dbConnect(duckdb::duckdb(), ":memory:")
dbWriteTable(con, "panel", as.data.frame(workshop_panel), overwrite = TRUE)

panel_db <- tbl(con, "panel")
```

A real job points the connection at a file on disk rather than `":memory:"`, so
the database, the temporary tables, and the intermediate joins all live on disk
and never have to fit in RAM at once. Nothing else about the code changes.

## The same verbs, unchanged

Here are the three staged strategies, identical to the staged-search article.

```{r strategies}
exact <- exact_strategy(
  workshop ~ normalize_text() + word_tokens(min_nchar = 3),
  block_by = c("postcode_area", "trade")
)

fuzzy <- search_strategy(
  workshop   ~ normalize_text() + word_tokens(min_nchar = 3),
  proprietor ~ normalize_text() + word_tokens(min_nchar = 2),
  block_by  = c("postcode_area", "trade"),
  threshold = 0.55
)

mover <- search_strategy(
  workshop ~ normalize_text() + word_tokens(min_nchar = 3),
  block_by     = list(block_on_tokens("workshop", max_df = 50, min_nchar = 4),
                      "trade"),
  rarity_scope = "global",
  threshold    = 0.6
)
```

Now run the staged search. This is the same `multi_stage_search()` call as the
in-memory article, with one addition: `control = duckdb_control()` says how the
run should be carved up on the database. We will look at that object in a moment;
the default is fine for a panel this size.

```{r run}
g <- multi_stage_search(
  panel_db, panel_db,
  base_id = "record_id", target_id = "record_id",
  list(exact = exact, fuzzy = fuzzy, mover = mover),
  self = TRUE, source_by = "year", collapse = "rep",
  control = duckdb_control(progress = FALSE)
)
```

The result is a lazy DuckDB table, not a data frame. Nothing has been pulled into
R yet. The entity grouping lives in the database until you ask for it with
`collect()`, which is the moment the plan runs and uses memory. On a corpus of
millions you `collect()` the grouping (one row per record, small) rather than the
intermediate joins, so what comes back to R stays modest no matter how big the
input was.

```{r collect}
grouping <- collect(g)
head(grouping, 4)
```

The columns are exactly the ones the in-memory backend returns: `entity` is the
recovered identity, `source` and `covered_sources` carry the year and the
trajectory length, `stage` records which pass attached the record. From here it is
an ordinary collected table, so score it exactly as the in-memory article does, by
checking how many true workshops land in a single entity.

```{r score}
intact <- as.data.table(grouping)[, .(record_id = id, entity)] |>
  merge(as.data.table(workshop_panel)[, .(record_id, true_entity)],
        by = "record_id")
intact[, .(n_entities = uniqueN(entity)), by = true_entity][
  , .(intact = sum(n_entities == 1), total = .N)]
```

This is the result that matters most in the whole article: it is **identical** to
the in-memory run, down to the per-stage counts. The backend changed where the
work happened, not what it found. The diagnostic verbs follow the table too, so
`compare_stages()` reads the DuckDB result with no special handling.

```{r compare}
compare_stages(g, base = panel_db, target = panel_db)
```

When you are finished, close the connection. A file-backed database also flushes
to disk here.

```{r disconnect}
dbDisconnect(con, shutdown = TRUE)
```

## Controlling how the run is carved up

On `r nrow(workshop_panel)` rows the defaults are invisible. On 50 million they are the difference
between a run that finishes and one that runs the machine out of memory. That is
what `duckdb_control()` is for. It is worth being precise about what it does and
does not touch.

The strategy decides **what matches**: which columns, how they are tokenized, how
rare a shared token has to be, where the threshold sits. The control decides
**how the run executes**: how the table is broken into pieces, how big those
pieces are, what happens when one of them fails. These are deliberately separate.
The same strategy needs heavy carving at 50 million rows and none at a thousand,
so how the work is split can never be a matching decision. You pass the control
to the same verbs, as `control =`.

```{r control}
ctrl <- duckdb_control(
  target_batch_size = 5e5,                 # rows per piece; NULL auto-tunes from memory
  chunk_by          = NULL,                 # key the scoring join splits on; NULL auto-derives
  chunk_strategy    = "block_consolidated", # how whole blocks are packed into pieces
  on_error          = "skip"                # what to do if one piece fails
)
```

Three ideas are worth understanding before you tune anything.

**Splitting is safe in two different ways, and the backend knows the difference.**
Turning text into tokens happens one row at a time, so that step can be split
anywhere. Finding matching pairs is different: a pair can only form between two
records in the same block, so a block can never be split across pieces or the
pairs that cross the split would simply vanish. The backend respects that
automatically. It packs whole blocks into each scoring piece and never cuts one in
half. Two knobs steer that packing. `chunk_by` names the key the scoring join
splits on; left at `NULL` it derives a sensible one when the table is large and
leaves small tables in one piece. `chunk_strategy` chooses how whole blocks are
grouped into pieces: the default `"block_consolidated"` packs small blocks
together to keep the piece count down, while `"block_first"` isolates blocks and
`"even"` ignores block structure entirely.

**A bad block should not sink the whole run.** On a corpus this size there is
always one pathological block, a postcode crammed with boilerplate names, that
blows up into far more pairs than the rest. `on_error = "skip"` records that piece
and moves on, so a single bad block costs you that block, not the twelve hours
that came before it. `"retry"` gives it one more attempt with more conservative
settings first; `"stop"` re-raises, which is what you want while you are still
debugging a strategy.

**Progress reports itself on long jobs.** When a run is large enough to take real
time, the backend prints which piece it is on out of how many, so a multi-hour job
is never a silent prompt. You can force it on or off with `progress`.

## At scale: a real directory

The synthetic panel above is the mechanism. Here is what the same mechanism did on
the corpus it was built for.

The motivating job was a national business directory, published yearly for two
decades: roughly **51 million listings**, the same firms reappearing year after
year under names and addresses that drift, get re-spelled, and relocate. The goal
was one panel that follows each firm through time, the same goal as the small
example, three orders of magnitude larger.

The shape of the work was exactly the staged search you just ran, with the
database doing the carrying:

- Block by region and industry, so a firm is only ever compared against plausible
  neighbours rather than the whole country.
- Deduplicate within each year first, then link across years in increasingly
  tolerant passes, collapsing each firm as it is found so later passes search a
  smaller space. This is the `collapse = "rep"` bridge from the small example,
  earning its keep across twenty years of drift instead of five.
- Fill short gaps in each firm's trajectory afterward, so a firm missing from a
  single year's directory is not mistaken for two separate firms.

That pipeline resolved the 51 million listings into about **7.5 million distinct
firms**, with roughly **2% of panel rows** reconstructed as short within-firm gaps.
The same `duckdb_control()` failure isolation above mattered in practice: on a
corpus that size a handful of dense blocks misbehave, and skipping them kept the
run moving instead of forcing a restart from the top.

The lesson is the one this article opened with. The strategy you debug on a
thousand rows in memory is the strategy you run on fifty million on disk. Only the
control around it changes.

## Where to look next

- The strategies used here, and the reasoning behind the exact, fuzzy, and mover
  passes, are the subject of [matching across years and sources](staged.html).
- The individual matching tools (containment, token blocking, global rarity,
  phonetic encoders) get their full treatment in the
  [features article](features.html).
- Deciding how strict to be, and what a wrong or missed link costs, is the
  [calibration article](calibration.html). It runs on a database backend too.
