---
title: "Embedding-based matching"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Embedding-based matching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# This article calls an external embedding provider, so its chunks are not
# evaluated when the site is built. The code is exactly what you would run.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

Every other strategy in joinery matches on surface form: shared tokens, shared
sounds, shared subsets. That works until two records mean the same thing while
sharing none of the same words. "J. Pollard, Cabinetmaker" and "Pollard Bespoke
Furniture" are the same workshop to a human and a total mismatch to a token
scorer. Embeddings are how you match on meaning instead of spelling.

This article shows when that helps, what it costs, and where it quietly fails. It
uses the same `workshop_register` / `workshop_listings` pair as the other
articles, so you can read the behaviour off the page.

## What an embedding is

An **embedding** turns a piece of text into a list of numbers (a vector) that
captures its meaning. A good embedding model places texts that mean similar
things close together in that number space and texts that mean different things
far apart. So once every record is a vector, "are these two records similar?"
becomes "are these two vectors close?", which is a single arithmetic operation
(cosine similarity) rather than a token comparison.

The useful part is that closeness survives a change of words. "Cabinetmaker" and
"bespoke furniture maker" land near each other even though they share no tokens,
because the model learned they mean roughly the same trade. That is the one thing
token and phonetic matching cannot do.

You do not train anything. You send each record's text to a pre-trained
embedding model and get its vector back. joinery does this through the
[tidyllm](https://edubruell.github.io/tidyllm/) package, which talks to embedding
providers (OpenAI, Mistral, Voyage, or a local model via Ollama). Which model you
use, and how to set it up, is covered in tidyllm's own guide:
[Embedding Models in tidyllm](https://edubruell.github.io/tidyllm/articles/tidyllm_embed.html).
Read that first if you have not used embeddings before; it explains the
providers, the API keys, and how to run a model locally for free.

One thing to know before you start: embeddings are powerful and need almost no
configuration, but they are not free. Turning text into vectors (**generation**)
calls an external model and is the expensive, slow step. On a few thousand
records that is seconds; on millions it is the cost that dominates everything.
The rest of this article is about getting the power without paying that bill
twice.

```{r load}
library(joinery)
library(tidyllm)
library(dplyr)
```

## The one-call matcher

You name the model and the columns to embed, and you are done. There is no
preparer pipeline, no rarity tuning, no blocking required. Here we use a local
Ollama model; swap in any provider tidyllm supports.

```{r strategy}
emb_model <- ollama(.model = "mxbai-embed-large")

st_embed <- embedding_strategy(
  columns         = c("workshop", "proprietor", "town"),
  embedding_model = emb_model,
  threshold       = 0.80
)
```

`embedding_strategy()` folds the three named columns into one text string per
record, embeds it, and scores a pair by the cosine similarity of their two
vectors. `search_candidates()` then returns every cross-table pair scoring at or
above the `threshold`.

```{r search}
m <- search_candidates(
  workshop_register, workshop_listings,
  base_id = "reg_no", target_id = "listing_id",
  strategy = st_embed
)
```

On the linkable listings this recovers around 88% of the true matches with zero
configuration: no token rules, no phonetic encoders, no `block_by`. The single
vector folds name, proprietor, and town into one signal and gets most of the way
on its own. This is what makes embeddings seductive, and it is a genuinely good
first cut on small, reasonably clean data.

## The bill is all in generation

It is worth seeing where the time actually goes, because it changes how you use
the tool. There are two cost axes, and they are about 100x apart.

```{r cost-probe}
# Generation: text -> vectors, one HTTP round-trip per batch to the model.
system.time({
  base_emb   <- compute_embeddings(workshop_register, "reg_no", st_embed)
  target_emb <- compute_embeddings(workshop_listings, "listing_id", st_embed)
})
#>    user  system elapsed
#>   0.6     0.1    40.1        # ~1,950 records, ~49 records/sec

# Retrieval: scoring every pair from vectors you already have. One matmul.
system.time(
  score_embeddings(base_emb, target_emb, st_embed)
)
#>    user  system elapsed
#>   0.36    0.02   0.40        # ~940k pairs
```

Generation took about 40 seconds; scoring every pair took under half a second.
So when people say "embeddings are expensive" they mean **generation** is
expensive, and it is expensive because it is an external call, not because of
anything joinery does. Everything downstream, the scoring, the thresholding, the
sweep below, is effectively free once the vectors exist.

This is the whole reason the next two sections matter. The lever that controls
cost is never "score faster", it is "embed fewer rows, and never embed the same
row twice".

## Embed once, reuse forever

joinery remembers vectors it has already computed. The first call to a verb that
needs embeddings pays the generation cost; later calls reuse the vectors instead
of re-embedding. A multi-stage run, a re-run with a different threshold, or a
second verb on the same data all reuse silently.

```{r reuse}
# First call: pays generation.
search_candidates(workshop_register, workshop_listings,
                  "reg_no", "listing_id", st_embed)

# Same data again at a stricter threshold: re-embeds nothing, returns instantly.
search_candidates(workshop_register, workshop_listings,
                  "reg_no", "listing_id",
                  embedding_strategy(
                    columns = c("workshop", "proprietor", "town"),
                    embedding_model = emb_model,
                    threshold = 0.90
                  ))
```

The cache is keyed by the model and the record's text, so it does the safe thing
automatically: change a record's text and only that record re-embeds; switch to a
different model and nothing stale is reused.

By default the cache lives for the R session. To keep vectors across sessions, so
tomorrow's run does not re-pay today's bill, point it at a directory:

```{r cache-disk}
options(joinery.embedding_cache_dir = "~/.cache/joinery-embeddings")
```

Now every vector is also written to disk and reloaded on the next run. To force a
clean re-embed or reclaim memory in a long session, clear it:

```{r cache-clear}
clear_embedding_cache()             # session cache
clear_embedding_cache(disk = TRUE)  # session and disk
```

The DuckDB backend does the same thing its own way: it stores vectors in a real
column on the table and skips rows that already have one. Either way, the rule is
the same. Embed once.

## The ceiling: where meaning is too blunt

Embeddings get you most of the way at no effort, which invites the question of
how far you can push them. Walk the threshold up and watch precision and recall
trade off. This sweep is free once the vectors exist, so it costs nothing to look.

| threshold | accepted | precision | recall |
|-----------|----------|-----------|--------|
| 0.60      | 893      | 0.79      | 0.91   |
| 0.70      | 884      | 0.80      | 0.91   |
| 0.75      | 847      | 0.83      | 0.91   |
| 0.80      | 773      | 0.89      | 0.88   |
| 0.85      | 695      | 0.91      | 0.81   |
| 0.90      | 537      | 0.92      | 0.63   |
| 0.95      | 266      | 0.94      | 0.32   |

Read down the column: precision climbs to about 0.94 and then flattens. Past
about 0.85 each step up buys almost no precision and gives up recall fast. That
ceiling is not a tuning problem you can fix with a better threshold. It is
structural.

The records stuck under that ceiling are the genuine look-alikes: two unrelated
"Oakwood Joinery" workshops in different towns, a common proprietor surname, a
generic trade. The `workshop_listings` data plants these on purpose in its
`homonym_*` tiers. An embedding pulls these pairs close together because they
really do read alike, and no threshold can separate "alike because same business"
from "alike because similar name". The model compressed away the distinction.

This is exactly the distinction joinery's token and rarity machinery is built to
make. A rare shared token ("Oakwood" appearing in only two records) is strong
evidence; a common one ("Joinery" in hundreds) is weak. Rarity scoring reads that
difference directly, which is why a token strategy can hold two same-name
workshops apart where an embedding fuses them. Embeddings are a floor to build
on, not a ceiling to chase.

## Where embeddings belong: the residual

The honest place for embeddings is not as the primary matcher but as the last
stage. Cheap token and phonetic passes do the bulk of the work and cost almost
nothing, then embeddings mop up the residual, the records where the surface forms
genuinely disagree but the meaning matches. Because you only embed the small
residual, the generation bill stays small.

`multi_stage_search()` runs strategies in order, and each stage only sees what
the earlier stages left behind. Put the cheap, precise strategies first and the
embedding stage last:

```{r staged}
st_exact <- exact_strategy(
  workshop ~ normalize_text() + word_tokens(min_nchar = 3),
  containment            = "forward",
  min_containment_tokens = 3,
  block_by               = c("postcode_area", "trade")
)

st_fuzzy <- search_strategy(
  workshop ~ normalize_text() + word_tokens(min_nchar = 3),
  block_by  = c("postcode_area", "trade"),
  threshold = 0.7
)

m_staged <- multi_stage_search(
  workshop_register, workshop_listings,
  base_id = "reg_no", target_id = "listing_id",
  strategies = list(st_exact, st_fuzzy, st_embed)
)
```

The exact and fuzzy passes recover the typos, slogans, and reorderings for free.
The embedding stage runs only on the listings neither could place, which is the
small set where the words truly differ. That is the pattern to remember: tokens
for the bulk, embeddings for the remainder.

There is a catch worth stating plainly. The residual is the **hard** set. It is
enriched for exactly the near-twins and homonyms that sit under the precision
ceiling, so an embedding stage on the residual is working in its weakest region.
Do not trust raw top-scoring pairs there. Pair the embedding stage with a higher
threshold, and lean on the calibrated false-positive filter, which is built for
this and reads embedding features (`cosine_sim`, the per-record vector norms)
directly. See [calibrating a false-positive filter](calibration.html) for how to
fit one. Calibration is the precision answer that the threshold alone cannot give
you.

## Scaling: blocking and the cost wall

Everything above runs in seconds because the data is small. Two things change as
the data grows.

Generation cost grows with the number of records, and it is the wall you hit
first. The defence is the same one this article has built toward: only embed what
you must. Stage embeddings last so you embed the residual, not the corpus, and
keep the disk cache on so a row is embedded once across all your runs.

Scoring cost grows with the number of *pairs*, which is the product of the two
table sizes. Below a few hundred thousand records the all-pairs scan is fine, as
the sub-second scoring figure above shows. Beyond that, give the embedding strategy a
`block_by` so it only scores pairs within the same block, the same lever the
token strategies use:

```{r blocked}
st_blocked <- embedding_strategy(
  columns         = c("workshop", "proprietor", "town"),
  embedding_model = emb_model,
  threshold       = 0.80,
  block_by        = c("postcode_area", "trade")
)
```

Within a block the scan is exact: every pair is compared, nothing is missed.
joinery does not yet use an approximate nearest-neighbour index, so for a very
large corpus that you cannot block sensibly, the right tool is a dedicated vector
store rather than joinery's in-database scan. For the blocked, residual-sized
problems embeddings are actually good at, the exact scan is both correct and fast
enough.

## Where to look next

- **Cutting false positives**: the residual-rescue pattern needs a precision
  filter. [Calibrating a false-positive filter](calibration.html) fits one from
  labelled pairs and uses the embedding similarity as a feature.
- **The token and phonetic strategies** that should run before the embedding
  stage: [beyond the basics: fuzzy and exact strategies](features.html).
- **Choosing and configuring a model**:
  [Embedding Models in tidyllm](https://edubruell.github.io/tidyllm/articles/tidyllm_embed.html).
