---
title: "Beyond the Basics: Fuzzy and Exact Strategies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beyond the Basics: Fuzzy and Exact Strategies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The Getting Started guide used one tool: a `search_strategy()` that scores every
candidate pair by how much rare token mass the two records share. That fuzzy
score is forgiving, which is what you want for typos and reorderings, but
forgiveness costs precision and compute. Two things round it out:

- an **exact strategy** that matches by token-set identity or containment. It is
  cheap and certain, and it handles the common case where one side simply
  appends extra words.
- **staging**, where you run several strategies in order with
  `multi_stage_search()`: a strict, cheap pass first, then looser passes that
  only see what the earlier passes left behind.

This article builds both on the `workshop_register` / `workshop_listings` pair.
Every feature has a planted tier in the data, so switching it on produces a
number you can read off the page.

```{r load}
library(joinery)
library(dplyr)
```

## The data

`workshop_register` is the clean base table: 982 entries from a fictional Guild
of Master Craftsmen roll. `workshop_listings` is the messier directory side: 844
listings, 729 of which refer to a known register entry (the `actual_link`
column). The matchable columns share names across both tables, so one strategy
formula serves both sides. Block on `(postcode_area, trade)`.

```{r glimpse}
glimpse(workshop_register)
```

The listings come in labelled tiers (`gen_tier`), each exercising one challenge:

| Tier | n | Challenge |
|---|---|---|
| `clean` | 290 | Minimal distortion. |
| `slogan` | 119 | Directory appends a marketing slogan after the name. |
| `variant` | 96 | Abbreviations, punctuation, word reorderings. |
| `mover` | 86 | Workshop relocated; listing is in a different `postcode_area`. |
| `phonetic` | 21 | Name spelled by ear (Pennell / Penell). |
| `hub_member` / `hub_trap` | 42 | A shared venue name baits over-linking. |
| `category_trap` | 8 | A bare generic trade name, another over-linking bait. |
| `homonym_*` | 84 | Different workshops with similar names. |
| `new` | 98 | No match in the register. |

The trap and homonym tiers carry no `actual_link` (there is nothing correct for
them to match); they are here to test precision, and we return to them at the
end.

### One way to measure

Every verb in this article returns the same thing: an entity table, one row per
record, with an `entity` id that groups records the matcher believes are the
same workshop. A listing is **recovered** when it lands in the same entity as the
register row named in its `actual_link`. That one rule scores every strategy
below, so we write it once.

```{r recall-helper}
entity_recall <- function(ms, tiers = c("clean", "slogan", "variant", "mover", "phonetic")) {
  ent   <- ms |> select(id, entity)
  truth <- workshop_listings |>
    filter(!is.na(actual_link)) |>
    select(listing_id, actual_link, gen_tier)

  truth |>
    left_join(ent, by = c("listing_id" = "id")) |>
    left_join(ent, by = c("actual_link" = "id"), suffix = c("_listing", "_register")) |>
    mutate(linked = !is.na(entity_listing) & entity_listing == entity_register) |>
    filter(gen_tier %in% tiers) |>
    group_by(gen_tier) |>
    summarise(recovered = sum(linked), total = n(), .groups = "drop") |>
    arrange(match(gen_tier, tiers))
}
```

`multi_stage_search()` produces that entity table even for a single strategy, so
we run everything through it and the same helper scores all of them.

## Exact matching: token sets, not scores

Many directory listings differ from the register in a way that has nothing to do
with spelling. They just say more. A slogan listing looks like this:

```{r slogan-example}
workshop_listings |>
  filter(gen_tier == "slogan") |>
  select(listing_id, workshop) |>
  slice(1:3)
```

The register name is "Pollard Joinery"; the listing is "Pollard Joinery" plus a
tagline. The two token sets are not equal, but the register's tokens are a
**subset** of the listing's. That relationship is exactly what an exact strategy
with containment is for.

```{r exact-strategy}
st_exact <- exact_strategy(
  workshop ~ normalize_text() + word_tokens(min_nchar = 3),
  containment = "forward",
  block_by    = c("postcode_area", "trade")
)
```

`containment` has three settings:

- `"off"`: the two token sets must be **equal**. The strictest match.
- `"forward"`: the base record's tokens must be a **subset** of the target's.
  This catches the slogan case, where the directory adds words.
- `"bidirectional"`: either side may be the subset of the other.

Every pair an exact strategy returns scores `1.0`. There is no rarity weighting
and no threshold to tune: the token sets either stand in the required
relationship or they do not. That makes it cheap, and on the slogan tier it is
also complete.

```{r exact-run}
m_exact <- multi_stage_search(
  workshop_register, workshop_listings,
  base_id = "reg_no", target_id = "listing_id",
  strategies = list(st_exact)
)

entity_recall(m_exact)
```

All 119 slogan listings recovered, plus most of the clean tier, at no scoring
cost. Movers and phonetic variants stay at zero, for two different reasons: a
phonetic spelling is not a subset of anything (containment has no notion of
"close"), and a mover never shares a `postcode_area` block with its register
entry in the first place. Those are the two jobs we stage on top of this later.

## When containment over-links: the shared venue

Containment is powerful precisely because it ignores extra words, and that is
also how it gets you into trouble. Consider a shared workshop building that rents
benches to several independent businesses, and that is itself a guild member:

```{r mall-register}
workshop_register |>
  filter(workshop == "Trinity Workshops") |>
  select(reg_no, workshop, trade, postcode_area)
```

"Trinity Workshops" is two tokens. Every business at that address lists itself as
"<their name>, Trinity Workshops":

```{r mall-listings}
workshop_listings |>
  filter(grepl("Trinity Workshops", workshop)) |>
  select(listing_id, workshop, actual_link)
```

Under forward containment the venue's two tokens are a subset of every one of
those listings. So the venue's register row links to all of them, and connected
components then fuse the unrelated workshops behind those listings into a single
entity:

```{r mall-trap}
st_loose <- exact_strategy(
  workshop ~ normalize_text() + word_tokens(min_nchar = 3),
  containment            = "forward",
  min_containment_tokens = 1,
  block_by               = c("postcode_area", "trade")
)

m_loose <- multi_stage_search(
  workshop_register, workshop_listings,
  base_id = "reg_no", target_id = "listing_id",
  strategies = list(st_loose)
)

# find the entity the venue (GMC-V0001) landed in, then list everyone in it
trap_entity <- m_loose$entity[m_loose$id == "GMC-V0001"][1]
m_loose |>
  filter(entity == trap_entity) |>
  select(id, rep, score)
```

Three distinct register entries (`GMC-00413`, `GMC-00500`, `GMC-00522`) are now
one entity, joined only because they happen to share a building. This is a real
pattern: it is the shopping-mall problem that motivated the containment guards in
the first place.

### The two guards

An exact strategy has no threshold, so it cannot be tuned the way a fuzzy score
can. Instead it carries two guards that decide which proper-containment links are
allowed to stand:

- **`min_containment_tokens`** is a cardinality floor. A proper-subset link is
  only kept when the contained (base) record has at least this many tokens. A
  two-token venue name cannot reach a floor of three, so it can no longer absorb
  longer listings.
- **`min_base_rarity`** is a rarity-mass floor. It drops links where the
  contained record's tokens are individually too common to carry identifying
  weight, even if there are several of them.

Setting the cardinality floor to three closes the trap:

```{r mall-fixed}
st_guarded <- exact_strategy(
  workshop ~ normalize_text() + word_tokens(min_nchar = 3),
  containment            = "forward",
  min_containment_tokens = 3,
  block_by               = c("postcode_area", "trade")
)

m_guarded <- multi_stage_search(
  workshop_register, workshop_listings,
  base_id = "reg_no", target_id = "listing_id",
  strategies = list(st_guarded)
)

fixed_entity <- m_guarded$entity[m_guarded$id == "GMC-V0001"][1]
m_guarded |>
  filter(entity == fixed_entity) |>
  select(id, rep, score)
```

The venue now keeps only its own true match: the bare "Trinity Workshops"
listing, an equality match that the cardinality guard never touches (the floor
gates proper-subset links, not identical sets). The three real workshops are
free again.

### The guard is not free

That safety has a price. A floor of three also blocks legitimate subset links
where the register name is short:

```{r guard-cost}
entity_recall(m_guarded, tiers = "slogan")
```

Slogan recovery drops from 119 to 106. The thirteen we lost are workshops whose
register name is only two tokens ("Pollard Joinery"), so their genuine slogan
listings can no longer clear the cardinality floor either. The guard cannot tell
a two-token shop from a two-token mall. Tightening exact matching to keep out the
mall necessarily keeps out those short real names too. The next section is how
you get them back.

## Staging: strict first, then forgiving

The exact pass is cheap and precise but rigid; it drops the short slogans and
never had a chance at movers or phonetics. A fuzzy `search_strategy()` is the
opposite: tolerant, but more expensive and lower precision. You do not have to
choose. `multi_stage_search()` runs a list of strategies in order, and each
later stage only sees the records the earlier stages failed to match.

Put the guarded exact strategy first and a plain fuzzy strategy second:

```{r stage-exact-fuzzy}
st_fuzzy <- search_strategy(
  workshop ~ normalize_text() + word_tokens(min_nchar = 3),
  block_by  = c("postcode_area", "trade"),
  threshold = 0.7
)

m_staged <- multi_stage_search(
  workshop_register, workshop_listings,
  base_id = "reg_no", target_id = "listing_id",
  strategies = list(st_guarded, st_fuzzy)
)

entity_recall(m_staged)
```

Slogan recovery is back to 119: the strict exact pass took the long names safely,
and the fuzzy pass picked up the thirteen short ones it had skipped, along with
more of the variant tier. You kept the precision of exact matching on the bulk of
the work and spent the expensive fuzzy scorer only on the residual.

## A stage that follows names across geography

Movers are still at zero. No mover listing shares a `postcode_area` with its
register entry, and both stages so far block on `postcode_area`, so the two sides
never meet.

```{r mover-example}
workshop_listings |>
  filter(gen_tier == "mover") |>
  select(listing_id, workshop, postcode_area, actual_link) |>
  slice(1:3)
```

The fix is to block on the name itself instead of on geography. Two records
co-block when they share a rare workshop-name token, so a distinctive name
travels with the business wherever it moved.

```{r rf-strategy}
st_region_free <- search_strategy(
  workshop ~ normalize_text() + word_tokens(min_nchar = 3),
  block_by     = list("trade", block_on_tokens("workshop", max_df = 50)),
  rarity_scope = "global",
  threshold    = 0.7
)
```

Two arguments carry this:

- `block_on_tokens("workshop", max_df = 50)` makes each record block on its own
  rare name tokens. The `max_df = 50` cap excludes boilerplate like "Joinery" or
  "Ltd" that appears in too many entries to be a useful block key.
- `rarity_scope = "global"` measures token rarity across the whole corpus rather
  than within a block, so a nationally distinctive name reads as rare even where
  several similar names cluster locally.

Before setting the cap, look at what the hottest tokens are with
`rarity_distribution()`:

```{r rarity}
rarity_distribution(workshop_register, id = "reg_no", strategy = st_region_free)
```

Without the cap, a common descriptor would pull a large share of the corpus into
one block and make the overlap join both slow and noisy. The cap is the lever
that keeps token blocking affordable.

## A stage that matches names by sound

The phonetic tier spells surnames by ear:

```{r phonetic-example}
phon <- workshop_listings |>
  filter(gen_tier == "phonetic") |>
  select(listing_id, workshop, actual_link)

phon |>
  left_join(select(workshop_register, reg_no, register = workshop),
            by = c("actual_link" = "reg_no")) |>
  select(listing = workshop, register) |>
  head(5)
```

"Penell" and "Pennell" are different token strings, so neither exact nor token
fuzzy matching connects them, but they share a Cologne phonetic code. A stage
that drops short tokens (so codes for "Ltd" or "LLP" do not flood the match) and
then encodes what remains will catch them:

```{r phonetic-stage}
st_phonetic <- search_strategy(
  workshop ~ normalize_text() + word_tokens() + drop_short_tokens(min_nchar = 4) + as_cologne(),
  block_by  = c("postcode_area", "trade"),
  threshold = 0.7
)
```

## The full pipeline

Stack all three in order of how much you trust them: exact first, then
region-free fuzzy for the bulk and the movers, then phonetics for the residual
sound-alikes.

```{r full-pipeline}
m_full <- multi_stage_search(
  workshop_register, workshop_listings,
  base_id = "reg_no", target_id = "listing_id",
  strategies = list(st_guarded, st_region_free, st_phonetic)
)

entity_recall(m_full)
```

Every tier is now covered: the movers came in through token blocking, the
phonetic names through the Cologne stage, and the slogan and variant tiers
through the earlier passes. `compare_stages()` shows what each stage actually
contributed:

```{r compare-stages}
stages <- compare_stages(m_full)
stages
```

Each stage adds matches the earlier ones could not reach, and it only ever works
on their leftovers, so the strict, cheap exact pass carries most of the load and
the expensive stages run on a shrinking residual. The default plot draws that as
a cumulative coverage curve:

```{r compare-stages-plot, fig.alt = "Cumulative base records matched, rising and flattening across the three stages.", fig.width = 7, fig.height = 4.5}
plot(stages)
```

The curve climbs steeply at the exact pass, then flattens: each later stage costs
more per record recovered, which is exactly why you order them strict-to-loose.

## Where to look next

- **Large datasets**: the same strategies run on a DuckDB backend with no formula
  changes. See *Working at scale with DuckDB*.
- **Names that describe the same thing in different words**: when two records
  share no tokens and no sounds (synonyms, free text, different languages), token
  and phonetic matching both fail. `embedding_strategy()` matches on vector
  similarity instead. See *Embedding-based matching*.
- **Separating look-alikes that are not the same business**: the homonym tiers
  need more than a threshold. A calibrated post-match filter learns the
  difference from labelled pairs. See *Calibrating a false-positive filter*.
