---
title: "Matching across years and sources"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Matching across years and sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A business register is usually a yearly snapshot: a list of the workshops trading
in some area this year, then another list next year, and so on. Stack several
years of those snapshots together and you can do something a single year cannot.
You can follow each workshop through time and read off when it opened, when it
closed, and how long it lasted.

The obstacle is that a workshop is not recorded the same way every year. It is
written one way in 2019 and a little differently in 2021; one relocates to a new
part of town; one trades under a name spelled by ear. Pool all the years into one
table and the job is no longer "find the duplicates in a list". It is "decide
which rows, across all the years, are the same workshop", so that each workshop
becomes one trajectory rather than a scatter of unconnected yearly entries.

This article does that on a small multi-year panel of woodworking workshops. It
pools the years, links across them in ordered passes, and turns the result into
one stable identity per workshop. The payoff it builds toward: once you have those
trajectories you can count how many workshops open and close each year, and the
matching choices you make upstream decide whether those counts come out right.

```{r load, message = FALSE}
library(joinery)
library(dplyr)
```

## The data

`workshop_panel` is the same universe of woodworking workshops as the rest of the
examples, observed across five years. Each row is one workshop in one year. The
`true_entity` column is the answer key: every year-row of one workshop shares it,
and our job is to recover that grouping from the names alone. The `change_tier`
column records what makes each trajectory hard.

```{r data}
library(data.table)
panel <- as.data.table(workshop_panel)

c(rows = nrow(panel),
  workshops = uniqueN(panel$true_entity),
  years = uniqueN(panel$year))

panel[, .N, by = change_tier]
```

Most workshops just carry light year-to-year noise (`stable`). The hard minority
is what the passes below are built for: `name_drift` workshops change their name
part-way through, `mover` workshops relocate to a different postcode area, and
`phonetic` ones have a spelled-by-ear twin.

### One way to measure

Every run below returns the same thing: an entity table, one row per record, with
an `entity` id that groups the records the matcher believes are one workshop. A
trajectory is **fully recovered** when all of a true workshop's year-rows land in
a single entity. That one rule scores every approach, so we write it once.

```{r helper}
trajectories_intact <- function(grouping) {
  grouping |>
    select(record_id = id, entity) |>
    left_join(select(workshop_panel, record_id, true_entity),
              by = "record_id") |>
    group_by(true_entity) |>
    summarise(n_entities = n_distinct(entity), .groups = "drop") |>
    summarise(intact = sum(n_entities == 1), total = n())
}
```

There are 320 true workshops, so 320 is a perfect score.

## Linking the pooled table to itself

When you link two different sources you search one against the other. Here there
is one pooled table, so you search it against itself. `multi_stage_search()` does
this with `self = TRUE`: it pools the rows, runs each strategy as a search pass,
and groups the accumulated links into entities at the end. Tell it which column
records where a row came from with `source_by`, here the `year`, so the result can
report how many years each workshop spans.

Start with a single forgiving pass and see how far one strategy gets.

```{r single}
fuzzy <- search_strategy(
  workshop   ~ normalize_text() + word_tokens(min_nchar = 3),
  proprietor ~ normalize_text() + word_tokens(min_nchar = 2),
  block_by  = c("postcode_area", "trade"),
  threshold = 0.55
)

g1 <- multi_stage_search(
  panel, panel,
  base_id = "record_id", target_id = "record_id",
  list(fuzzy = fuzzy),
  self = TRUE, source_by = "year"
)

trajectories_intact(g1)
```

One pass already recovers most of the panel. What it misses are the trajectories
where the name moved too far in one step for a single threshold to bridge.

## Staging: a strict pass, then looser ones

The fix is to run several strategies in order, strictest first. An exact pass
matches only workshops whose name tokens are identical, which is cheap and almost
never wrong, and it clears the easy bulk out of the way. Then a fuzzy pass works
on what is left.

The important option is `collapse`. Between stages, `collapse = "rep"` takes every
group found so far and collapses it to a single representative row that stays
searchable. A drifting name is then linked one short step at a time: 2019 links to
2020, that pair collapses to one representative, and the representative is what the
next pass compares against 2021. Without the collapse, the looser pass has to
bridge the whole drift in one jump.

```{r staged}
exact <- exact_strategy(
  workshop ~ normalize_text() + word_tokens(min_nchar = 3),
  block_by = c("postcode_area", "trade")
)

g2 <- multi_stage_search(
  panel, panel,
  base_id = "record_id", target_id = "record_id",
  list(exact = exact, fuzzy = fuzzy),
  self = TRUE, source_by = "year", collapse = "rep"
)

trajectories_intact(g2)
```

To see what the collapse buys, run the same two stages with `collapse = "none"`,
which carries only the still-unmatched records forward and never builds a bridge:

```{r collapse-none}
g_none <- multi_stage_search(
  panel, panel,
  base_id = "record_id", target_id = "record_id",
  list(exact = exact, fuzzy = fuzzy),
  self = TRUE, source_by = "year", collapse = "none"
)

trajectories_intact(g_none)
```

Many more trajectories shatter without the carried representative. The drift
chains break into pieces because no single pass ever sees the two ends close
enough together. Collapse-and-continue is what holds a slowly drifting name in one
piece.

## A pass for the movers

The exact and fuzzy passes both block on `(postcode_area, trade)`: they only ever
compare workshops in the same area. That is fast and right for most of the panel,
but it can never link a workshop to its own later self once it has relocated,
because the two rows are in different areas. Those movers need a pass that does not
block on area at all.

Block instead on a rare word from the workshop name. Rarity here just means how
few records carry a token: a distinctive surname is rare, the word "joinery" is
common. A rare name token is shared by a workshop and its relocated self but by
almost nothing else, so it co-blocks the two rows wherever they sit. Measure that
rarity across the whole pool rather than within a block, so a distinctive token
reads as strong evidence no matter which area it appears in.

```{r mover}
mover <- search_strategy(
  workshop ~ normalize_text() + word_tokens(min_nchar = 3),
  block_by     = list(block_on_tokens("workshop", max_df = 50, min_nchar = 4),
                      "trade"),
  rarity_scope = "global",
  threshold    = 0.6
)

g3 <- multi_stage_search(
  panel, panel,
  base_id = "record_id", target_id = "record_id",
  list(exact = exact, fuzzy = fuzzy, mover = mover),
  self = TRUE, source_by = "year", collapse = "rep"
)

trajectories_intact(g3)
```

That recovers nearly all of the 320 trajectories. The three passes share the work
in a readable way: the exact pass links the clean rows, the fuzzy pass picks up the
drifters, and the mover pass reaches across the relocations the area block could
not.

```{r stage-table}
as.data.table(g3)[, .N, by = stage]
```

This table counts *records*, one per workshop-year, not trajectories. The `NA` row
is the workshops that never linked to anything: a workshop seen in only one year,
or one no pass could attach. Each is its own single-record entity.

## Reading the result

The entity table has one row per record and a handful of columns worth knowing:

```{r grouping-cols}
head(g3)
```

- `entity` is the recovered identity. Records that share it are one workshop.
- `rep` is the representative record chosen for the entity.
- `source` and `covered_sources` come from `source_by`. Here `source` is the row's
  year and `covered_sources` is how many distinct years the whole entity spans, the
  trajectory length.
- `n_in_entity` is how many records the entity holds; `stage` is the pass that
  attached this record.

Every link the passes found, with the stage and direction of each, is kept as a
ledger you can pull off the result. It is the audit trail behind the grouping:

```{r ledger}
ledger <- attr(g3, "ledger")
head(ledger[, c("from", "to", "stage", "source_from", "source_to",
                "within_source")])
```

`within_source` is `FALSE` for a cross-year link and `TRUE` for a within-year one,
so you can tell at a glance which links actually do the work of joining a
trajectory across time.

To see how the passes divide the labour, `compare_stages()` reports what each one
added:

```{r compare}
compare_stages(g3, base = panel, target = panel)
```

Read the "marginal coverage" block: it shows how many records each pass linked
that no earlier pass had reached, so you can see the exact front doing the bulk of
the work and the later passes earning their keep on the harder remainder.

## The payoff: entry and exit

Now turn the trajectories into the numbers you actually wanted. For each recovered
workshop, the first and last year it appears is its entry and exit; before 2019 or
after 2023 we cannot see, so a workshop first seen in 2019 may be older and one
still present in 2023 has not exited yet.

```{r entry-exit}
span <- as.data.table(g3)[, .(record_id = id, entity)] |>
  merge(panel[, .(record_id, year)], by = "record_id")
span <- span[, .(first = min(year), last = max(year)), by = entity]

c(births_after_2019 = span[first > 2019, .N],
  exits_before_2023 = span[last < 2023, .N])
```

Compare those to the truth, computed the same way on `true_entity`:

```{r entry-exit-truth}
truth <- panel[, .(first = min(year), last = max(year)), by = true_entity]

c(births_after_2019 = truth[first > 2019, .N],
  exits_before_2023 = truth[last < 2023, .N])
```

The counts are close but not equal, and they err in a specific direction. The few
trajectories the matcher still split show up as a workshop that "exits" when its
name drifts and a brand new one that "enters" the next year. Every missed link
manufactures one phantom exit and one phantom entry, so under-linking pushes both
counts up and makes the sector look more churny than it is.

That is the whole reason the matching threshold is not just a tuning detail when
you study entry and exit. Link too loosely and you fuse two different workshops,
erasing a real birth and a real death; link too strictly and you split one
workshop, inventing a birth and a death. The two mistakes bias churn in opposite
directions, and neither is safe by default. Choosing that operating point
deliberately, and checking what it costs, is its own job; the
[calibration article](calibration.html) is about exactly that decision.

## Where to look next

- The strategies used here (exact matching, fuzzy scoring, token blocking, and
  global rarity) get their full treatment in the
  [features article](features.html).
- Deciding how strict to be, and what a wrong or missed link costs your analysis,
  is the [calibration article](calibration.html).
- The same staged search runs on a database backend for panels too large for
  memory, with the verbs unchanged. That is the subject of the DuckDB article.
- For finding duplicates inside a single table rather than linking across years,
  reach for `multi_stage_dedup()` instead of `multi_stage_search()`.
