---
title: "Matching compound names with subword tokens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Matching compound names with subword tokens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Subword tokens need the optional sentencepiece package. Where it is missing,
# the chunks are shown but not run.
has_sentencepiece <- requireNamespace("sentencepiece", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_sentencepiece
)
data.table::setDTthreads(2)
```

You have a chamber register and a trade directory that list the same workshops.
The register writes each trade as one long word, `Holzbaugesellschaft Wagner
mbH`. The directory splits that word apart and puts the proprietor first,
`Wagner Holzbau KG`. Some directory lines went through a scanner that misread
letters on the way. Neither record holds a typo a spelling rule would fix, and
word tokens still cannot see the pair. `Holzbaugesellschaft` and `Holzbau` are
two different words, so the two records share only the surname.

Subword tokens fix that. You learn a vocabulary of frequent word pieces from
your own text, then split each record into those pieces. Two spellings of the
same name share most of their pieces even where they share no whole word, so
they meet and score like any other pair.

This article fits such a vocabulary on a small German register, measures what it
recovers against plain word tokens, and shows the two settings that decide
whether the result is any good.

Subword tokens use the `sentencepiece` package, which joinery loads only when
you ask for it:

```{r install, eval = FALSE}
install.packages("sentencepiece")
```

```{r library}
library(joinery)
```

## The data

Forty workshops, in one table to deduplicate. Twenty appear twice, once in the
register spelling and once in the directory spelling, and share at most the
proprietor's name. Ten more appear twice in near-identical spellings. Word
tokens already find those ten, so they are the control for what a subword run
gives up. Ten unrelated trades appear once and have no partner.

```{r data}
register <- c(
  "Holzbaugesellschaft Wagner mbH",  "Moebelwerkstatt Brandt",
  "Tischlereibetrieb Kaufmann",      "Fensterbaugesellschaft Lehmann",
  "Treppenbauwerkstatt Ritter",      "Parkettlegerbetrieb Sommer",
  "Innenausbaugesellschaft Naumann", "Kuechenmoebelfabrik Wendt",
  "Holzhandelsgesellschaft Gerlach", "Schreinereibetrieb Hoffmann",
  "Dachstuhlbaugesellschaft Kolbe",  "Bautischlereibetrieb Reuter",
  "Ladenbaugesellschaft Piotrowski", "Gartenmoebelwerkstatt Ebeling",
  "Massivholzmoebelfabrik Stein",    "Trockenbaugesellschaft Amrein",
  "Furnierhandelsgesellschaft Zorn", "Bootsbauwerkstatt Marquardt",
  "Saegewerksgesellschaft Uhlig",    "Restaurierungswerkstatt Thiele"
)

directory <- c(
  "Wagner Holzbau KG",              "Werkstatt fuer Moebel Brandt",
  "Kaufmann Tischlerei",            "Lehmann Fensterbau GmbH",
  "Ritter Treppenbau Werkstatt",    "Sommer Parkettleger",
  "Naumann Innenausbau GmbH",       "Wendt Kuechenmoebel Fabrik",
  "Gerlach Holzhandel",             "Hoffmann Schreinerei",
  "Kolbe Dachstuhlbau",             "Reuter Bautischlerei",
  "Piotrowski Ladenbau GmbH",       "Ebeling Gartenmoebel Werkstatt",
  "Stein Massivholzmoebel",         "Amrein Trockenbau GmbH",
  "Zorn Furnierhandel",             "Marquardt Bootsbau",
  "Uhlig Saegewerk",                "Thiele Restaurierung"
)

# The scanner misread every third directory line: rn as m, O as zero, ll as 11.
scanned <- seq(3, length(directory), by = 3)
directory[scanned] <- sub("ll", "11",
  sub("O", "0", sub("rn", "m", directory[scanned])))

# Ten workshops the directory copies almost verbatim.
plain <- c(
  "Tischlerei Baumgartner", "Schreinerei Kowalczyk",
  "Zimmerei Oberlaender",   "Drechslerei Feldmann",
  "Boettcherei Hanselmann", "Parkettstudio Vranken",
  "Holzwerkstatt Dallmann", "Moebelbau Steinhauser",
  "Treppenstudio Wendland", "Holzdesign Kretschmar"
)

# Ten trades that appear once and belong to nobody else.
others <- c(
  "Malerbetrieb Kessler",    "Dachdeckerei Vogt",
  "Elektrotechnik Sander",   "Sanitaertechnik Brunner",
  "Metallbau Hartwig",       "Steinmetzbetrieb Gruber",
  "Glaserei Wiedemann",      "Polsterei Nehring",
  "Zimmerei Achenbach",      "Schlosserei Berndt"
)

workshops <- rbind(
  data.frame(id = sprintf("R%02d", 1:30), firm = c(register, plain)),
  data.frame(id = sprintf("D%02d", 1:30), firm = c(directory, paste(plain, "GmbH"))),
  data.frame(id = sprintf("X%02d", 1:10), firm = others)
)

head(workshops, 3)
```

The ids carry the answer key: `R07` and `D07` are the same workshop, the `X`
rows belong to nobody. Two measures score every run below, so we write them
once. **Pair recall** is the share of the thirty known pairs whose two rows land
in one duplicate group. **Group purity** is the share of duplicate groups that
hold a single workshop. Anything under 1 means two different workshops were
joined.

```{r measures}
entities <- sprintf("%02d", 1:30)

pair_recall <- function(dups, which = entities) {
  if (nrow(dups) == 0) return(0)
  groups <- split(dups$id, dups$duplicate_group)
  mean(vapply(which, function(e) {
    any(vapply(groups, function(ids) all(paste0(c("R", "D"), e) %in% ids), logical(1)))
  }, logical(1)))
}

group_purity <- function(dups) {
  if (nrow(dups) == 0) return(NA_real_)
  groups <- split(sub("^[RD]", "", dups$id), dups$duplicate_group)
  mean(vapply(groups, function(e) length(unique(e)) == 1, logical(1)))
}
```

## Where word tokens stop

Start with the strategy you would reach for first: clean the text, split it on
whitespace, keep pairs that share enough rare word mass.

```{r word-run}
words <- search_strategy(
  firm ~ normalize_text() + word_tokens(),
  threshold = 0.5
)

word_dups <- detect_duplicates(workshops, id = "id", strategy = words)

c(recall = pair_recall(word_dups), purity = group_purity(word_dups))
```

Everything it finds is right, and it finds well under half. The ten
near-identical pairs come back, plus two compound pairs carried by the surname
alone. The other eighteen register rows do not meet their directory entry,
because the trade word the two records have in common sits buried inside a
longer word on one side.

## Learning a vocabulary

`find_subwords()` learns the pieces. Feed it the text as your strategy will see
it: the same cleaning steps that come before the tokenizer in the formula, here
`normalize_text()`. Fitting on raw text and applying to cleaned text gives worse
splits.

```{r fit}
corpus <- normalize_text(c(
  register, plain, others,
  "Holzbau Wagner", "Moebel Brandt", "Tischlerei Kaufmann",
  "Fensterbau Lehmann", "Treppenbau Ritter", "Parkettleger Sommer",
  "Innenausbau Naumann", "Kuechenmoebel Wendt", "Holzhandel Gerlach",
  "Schreinerei Hoffmann", "Dachstuhlbau Kolbe", "Bautischlerei Reuter",
  "Ladenbau Piotrowski", "Gartenmoebel Ebeling", "Massivholzmoebel Stein",
  "Trockenbau Amrein", "Furnierhandel Zorn", "Bootsbau Marquardt",
  "Saegewerk Uhlig", "Restaurierung Thiele",
  "Werkstatt", "Gesellschaft", "Betrieb", "Fabrik", "Handel", "GmbH", "KG"
))

sw <- find_subwords(corpus, vocab_size = 340)
sw
```

The corpus here is the names in the file plus the trade words on their own and
the words that mean company. A vocabulary learns pieces from the text it sees,
so a short word you want compounds cut at is worth feeding it separately.

One record shows what the model does to a compound:

```{r split-example}
subword_tokens(normalize_text("Holzbaugesellschaft Wagner mbH"), sw)
subword_tokens(normalize_text("Wagner Holzbau KG"), sw)
```

Both records now carry the piece `HOLZ`, and the directory record carries `BAU`
as well. That is enough shared mass for the pair to meet.

## What subword tokens find

Put the fitted model in the formula in place of `word_tokens()`. Nothing else
about the strategy changes. Pieces are ordinary tokens, each weighted by its
rarity, and the threshold means what it meant before.

```{r subword-run}
subwords <- search_strategy(
  firm ~ normalize_text() + subword_tokens(sw) + drop_short_tokens(min_nchar = 3),
  threshold = 0.5
)

sub_dups <- detect_duplicates(workshops, id = "id", strategy = subwords)

c(recall = pair_recall(sub_dups), purity = group_purity(sub_dups))
```

All thirty pairs come back, and each duplicate group holds a single workshop.
The ten pairs word tokens already had are still there:

```{r control}
pair_recall(sub_dups, which = sprintf("%02d", 21:30))
```

That formula also drops pieces under three characters. A learned vocabulary
contains one- and two-letter pieces, and those turn up in most records, so they
cost compute and blur the score. Dropping them is the cheapest improvement you
can make to a subword run.

## Two settings, pulling against each other

`vocab_size` decides how finely text is cut. A small vocabulary has few pieces
to work with, so it cuts words into short fragments that many workshops share. A
large one keeps frequent words whole and cuts only the rare ones, which brings
you back towards word tokens. The minimum piece length pulls the other way.

```{r grid}
grid <- expand.grid(vocab_size = c(200, 340, 400), min_nchar = c(2, 3))

result <- do.call(rbind, lapply(seq_len(nrow(grid)), function(i) {
  model <- find_subwords(corpus, vocab_size = grid$vocab_size[i])
  strat <- search_strategy(
    firm ~ normalize_text() + subword_tokens(model) +
      drop_short_tokens(min_nchar = grid$min_nchar[i]),
    threshold = 0.5
  )
  dups <- detect_duplicates(workshops, id = "id", strategy = strat)
  data.frame(
    vocab_size = grid$vocab_size[i],
    min_nchar  = grid$min_nchar[i],
    recall     = pair_recall(dups),
    purity     = group_purity(dups)
  )
}))

result
```

One corner of the grid is clean, 340 pieces with a three-character minimum. At
`vocab_size = 200` the recall holds and the purity falls, because the fragments
get short enough for unrelated workshops to share them. That is the failure to
watch for. A mixed duplicate group costs more to undo than a missed pair costs
to find. At 400 the vocabulary keeps the compounds whole again and pairs go
missing.

`rarity_distribution()` is how you judge a fitted vocabulary where you have no
ground truth to score against. It lists the pieces that turn up in the most
records. Here the list mixes whole words such as `WERKSTATT` with fragments
such as `EREI` and `MANN`, which unrelated workshops share. When fragments
crowd out the words at the top, the vocabulary cuts too finely.

```{r rarity}
rarity_distribution(workshops, id = "id", strategy = subwords)
```

On a corpus of millions of rows, fit on a sample. `find_subwords(x, sample_n =
2e5)` learns an equivalent vocabulary in a fraction of the time, because the
frequent pieces of a large corpus are already the frequent pieces of a large
sample of it.

## Keeping the model

A fitted vocabulary is a plain R object that carries the trained model as raw
bytes, so it saves and reloads like anything else and splits text the same way
every time.

```{r save, eval = FALSE}
saveRDS(sw, "workshop_subwords.rds")
sw <- readRDS("workshop_subwords.rds")
```

Save the model and rebuild the formula around it. A strategy formula carries
the environment it was written in, which makes it a much larger and less
predictable thing to put on disk. Rebuilding it around a saved model costs one
line and reproduces the same splits.

## Bringing your own tokenizer

`subword_tokens()` is a named shortcut for a more general step. `custom_tokens()`
takes any tokenizer and puts it in a strategy formula, so a rule you have written
yourself scores exactly like a built-in one.

The tokenizer can be a plain function that takes a character vector and returns a
list of token vectors, one per input record:

```{r custom, eval = TRUE}
split_runs <- function(text) strsplit(text, "[^a-z]+")

custom_tokens(c("holzbau wagner", "wagner-holzbau"), split_runs)
```

It can also be a fitted object with a `tokenize()` method. That is how a
tokenizer with learned state enters a strategy, and it is what a `Subword_Model`
is. `custom_tokens(text, sw)` and `subword_tokens(text, sw)` do the same work.

Whichever form you pass, the result needs one element per input record, each a
character vector of that record's tokens. A result of the wrong shape stops the
run with a clear message before any of it reaches the scorer.
